Universal Characterization of Physical Variability by \\Shannon's $-\sum_ip_i\,\textrm{ln}\,p_i$ Formula

### 2019-09-26 Articles/2019-ShannonEntropy/Source/ShannonEntropyIsVariability.tex
Incorporating comments from Julian and reviewing

\title{Universal Characterization of Variability by \\Shannon's $\sum_ip_i\,\textrm{ln}\,p_i$ Formula}

\author{
	Julian Barbour, 
	Gabriele Carcassi
}

@@

A world without any variety is barely conceivable. It certainly would not admit scientific investigation. One of us recalls but has been unable to relocate a sentence in one of Leibniz's many writings to the effect that reality \emph{is} variety. This is consistent with his view that any contingent thing is defined by a collection of attributes, which necessarily must be distinguishable
at some epistemic level. Paragraph 58 of Leibniz's \emph{Mondadology} asserts that the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''.

Without committing ourselves to Leibniz's overarching principle, our aim in this paper is to draw attention to the specific nature of the variety that can be observed in the universe and the unique suitability of Shannon's entropy (or information) formula $\sum_ip_i\,\textrm{ln}\,p_i$ for the characterization of the variety. Indeed, as we will show in the next section, opportunities for application of Shannon's formula are ubiquitous both on the earth and in the universe at large.

We also wish to comment on the various interpretations of Shannon's formula and attempt to answer this question: what does the formula represent? It is well known that when Shannon recognized the importance of his formula in communication theory he sought von Neumann's advice on the name he should give to the expression he had found. He was told to call it \emph{entropy}, first, because Botlzmann had used an identical formula to define entropy in statistical mechanics and, second, nobody understood entropy and that would provide protection from criticism [exact words to be found]. We believe von Neumann's advice was not totally helpful and has given rise to confusion. Indeed, it is often claimed that Shannon's formula is epistemic in nature and represents ignorance. There certainly are situations in which that is true but equally there are others in which the use of the formula is very appropriate and no ignorance is involved at all. Examples will be given in the next section.

This leads us to propose a different characterization of Shannon's formula that is both more general and more appropriate. Indeed, we will argue that the formula is a measure of the \emph{variability within a distribution of things of a given kind}. Moreover, the formula has a very wide range of applicability because the universe contains many different kinds of things that, through having common attributes, can be recognized as belonging together despite the fact that their attributes are expressed to different degrees or in different ways. To give a first example, the girls in a school can differ in the length and colour of their hair. More examples will follow in the next section.

@@

\vspace{2in}

The general setting is the following. We have a set of elements $E = \{e_\alpha\}_{\alpha=1}^N$. These can represent the galaxies in the Laniakea Supercluster, the animals in the Gal\'{a}pagos islands, the molecules in a box of gas given a particular macrostate, the different takes of spin measurement for a given quantum state, or the words in the complete works of Shakespeare. The choice of the particular set is driven by the interest and practices of scientists and their fields of study and therefore it is, in this sense, discretional. Yet, it must be objective in at least one respect: once a choice is made the actual elements are the same for everybody. Once we arbitrarily decided we will study the galaxies in the Laniakea Supercluster, it is a matter of fact that the Milky Way will be included. This also means that, whatever properties those elements will have, they will represent a matter of fact about that set. In turn, this will make some choices arguably more appropriate or interesting than others. For example, in taxonomy the set of all animals that have feathers may be more interesting than the set of all animals that can fly: the first includes only birds while the second includes most birds, a few mammals and a lot of insects.

Once a set of elements is chosen, we select a property or a set of properties we want to use to characterize the elements. That is, we have a set of possible values $Q=\{q_i\}_{i=1}^I$ and a map $q : E \to Q$ that associates a value for each element. These can represent the galaxy types, the genus of the animal, the possible states of the molecules, the possible values for spin or the spelling of words. This will give us a sequence $\{q(e_\alpha)\}_{\alpha=1}^N$ of the descriptions that are associates with each element in the set. Again, the choice of properties is ultimately discretional, grounded on what particular aspects of the elements we are studying in a given case. Yet, like before, once a choice is made the value for each element is objectively defined. Once we arbitrarily decided we are studying galaxy types it is a matter of fact that the Milky Way is a barred spiral galaxy. Again, one may find that some choices are arguably more appropriate or interesting than others.

Now that we have constructed our distribution, we want to construct an indicator $S$ that quantifies how much variability the elements exhibit within the distributions. That is, we want to quantify the degree of diversity that the values can have within the distribution. To that end, we want to define some suitable requirements for $S$.

We may be tempted to use standard statistical quantities, like the range or the variance, but this is not possible. First, if the values associated to the bins are non-numerical (e.g. types of galaxies, words), statistical quantities are not well defined. Second, permutation of the values does not have an impact on variability of the elements, while statistical quantities will in general be affected. This tells us that our indicator cannot depend on $Q$ itself, but only on the weights $p_i$. That is, we require that $H=H(p_i)$.\footnote{Note that for continuous quantities the weights $p_i$ are affected by the choice of $Q$. We will concentrate on this problem later.}

We  expect small changes in a distribution, small changes of the weights, to produce small changes in variability, therefore we require $H$ to be a continuous function with respect to the $p_i$. We also expect that if we have a uniform distribution $p_i = 1/I$, then greater $I$ should correspond to greater variability. Therefore, in the case of uniform distribution, we require $H$ to be monotonically increasing with $I$.

To sum up, we have the following three requirements:
\begin{enumerate}
    \item $H$ depends only on $p_i$ and it does so continuously
    \item If $p_i=1/I$ then $H$ is a monotonically increasing function of $I$
    \item Let $p_i$ and $q_j$ be the weights for two distribution respectively over $I$ and $J$ bins. Let $r_k$ be the distribution over $K=I+J-1$ bins constructed by expanding the $a$-th bin of the first distribution using the second distribution. More specifically, let $1 \leq a \leq I$, then $r_k = \{p_1, p_2, ..., p_{a-1}, p_{a}q_1, p_{a}q_2, ..., p_{a}q_J, p_{a+1}, ..., p_I \}$. Then $H(r_k) = H(p_i) + p_{a} H(q_j)$.
\end{enumerate}

The Shannon entropy, then, is a clear and important attribute of a distribution: it quantifies the variability of the elements within the distribution. It quantifies the variety of values within the distribution. It is true that the choice of the elements and binning are not fixed, it is true that the meaning of the weights depends on what the distribution is describing, and therefore that the precise meaning of this variability is context dependent. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope a probability and so on. And like a real number represents a quantity, transcending what is physically quantified, the Shannon entropy represents the variability of the elements in a distribution, transcending what the distribution physically signifies.

The Shannon entropy may represent knowledge in some cases, but not in general. Consider the following two cases:
\begin{enumerate}
    \item There is 50\% chance you won one million dollars in the lottery and 50\% chance you didn't;
    \item There is 50\% chance you won one million dollars in the lottery and 50\% chance you won half a million dollars.
\end{enumerate}
The distribution in both cases is the same, two bins 50\% chance each, and so is the Shannon entropy. Yet, you know more in the second case: you know you have won at least half a million dollars.

Unfortunately, entropy in general is often associated with vague characterizations as the ones presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending by the authors, which leads to confusion and misunderstanding.  The characterization of Shannon entropy we have given, on the other hand, applies to all cases, it is enough to understand the assumptions required to rederive it and therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos island, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability we find in its vocabulary? All these questions are interesting, all these questions are profound and all these questions are answered by a single concept: the Shannon entropy.

If the Shannon entropy is a measure of variability, why is it connected to information? What is it measuring and in what units?

We can gain insight with the following example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands. Suppose I pick a specific animal from the distribution and you want to know its species. Suppose the only way for you to get that information is to ask me a series of questions. Suppose each questions has two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

Not all questions will be able to extract the same amount of information. Some questions, like ``is it an animal?" would be redundant. Others, like ``is it an American Flamingo (Phoenicopterus ruber)?", would give us a lot of information in the positive case but little in the negative case. It should be clear, though, that there must be a limit to how few questions you must ask to get to the answer: a single yes/no question cannot be enough. It should, hopefully, be intuitive that to a greater variability within the distribution will correspond a greater number of questions you must ask. That is exactly what the Shannon entropy measures: the minimum average number of questions one has to ask to identify a value in the distribution. It gives us the number of questions for an ideal strategy for our game of Twenty Questions.

The Shannon entropy of the distribution, then, has nothing to do with the information the distribution itself holds. The distribution is not what is being encoded. The entropy is the information needed to go from the distribution, which is given, to an individual element. It's really the information gap from the population to an element. That is why some people say the entropy is ``lack of information", because it's the information the distribution does not have. But, again, this is just missing the point of what it is being described. If one is not interested in identifying elements there is no ``missing information".

As we saw in the previous sections, Shannon entropy is a very general concept that applies to any type of distribution, no matter what the problem domain. To understand its link to the thermodynamic/statistical mechanics entropies we need two answer two questions. First, what does the Shannon entropy represents in the context of statistical mechanics and why is it a state variable? Second, why does the Shannon entropy have the same properties of the thermodynamic entropy? That is, we have to understand why it increases during irreversible evolution, it is conserved during reversible evolution, it is maximized at equilibrium and it is additive for independent systems.

The first question is straight-forward to answer. The macroscopic variables of a system (e.g. volume, pression, energy) do not give a full description of its pars. If we fix a macrostate and then measured the state of all microscopic constituents, we would not find the same answer every time. The choice of a macrostate, then, identifies a set of possible microstates that are compatible with it. It is then valid to ask the question: what is the variability of microstates one can find within a given macrostates? This would be answered by the Shannon entropy, which will measure the variability in bits required to identify a microstate given the macrostate. This quantity is a property of the macrostate: it is a state variable.

For the others, suppose we have a volume of gas at a particular temperature and pressure in perfect equilibrium. While the microstate is constantly changing, the macrostate remains the same. Effectively, the equilibrium is simply permutating elements of the distribution. As we saw, permutations do not change the variability which has to remain constant. At equilibrium the Shannon entropy has to remain constant.

Suppose we compress the gas adiabatically and then let it re-expand adiabatically to the original state. As the system is isolated, the whole process is deterministic and reversible. An initial microstate has to map with a specific intermidiate and final microstate. Identifying an element at each stage is equivalent to identifying an element in any state. The variability within the distribution at each step, then, cannot change: the same number of bits, the same information, must be enough to identify initial, intermediate or finale microstates. Deterministic and reversible processes conserve Shannon entropy.

Now suppose we have a process that starts from an out-of-equilibrium state and ends in an equilibrium. The initial state will not be fully described by the standard macroscopic variables. In fact, there will be many out-of-equilibrium states that will lead to the same final equilibrium. In this sense the process is irreversible because knowing the final macroscopic variables will not be enough to reconstruct the initial macroscopic state. Different possible initial out-of-equilibrium states merge into a final equilibrium state.

If we assume, though, that the microstate evolves deterministically and reversibly, knowing the final microstate would allow us to reconstruct the initial microstate and what initial out-of-equilibrium macrostate would be compatible with it. As different out-of-equilibrium macrostate merge during the evolution, they merge the possible initial conditions that lead to that same macrostate. The variability of microstates within the initial out-of-equilibrium macrostate, then, must be smaller than the variability of the microstates withing the final equilibrium state. The amount of information needed to identify a microstate at equilibrium allows us to identify the initial microstate across multiple out-of-equilibrium states and therefore is greater than the information needed to identity a microstate in an out-of-equilibrium macrostate. Irreversible processes increase the Shannon entropy which, as we saw before, stabilizes at equilibrium.

Now suppose we have two independent volume of gasses. This means that picking the microstate of one system does not constrain the microstate of the other. That is, the distribution over microstates are statistically independent. In this case, the Shannon entropy is additive.

The Shannon entropy, then, has all the right properties to establish a conceptually meaningful connection with the thermodynamic/statistical mechanics entropies. Yet, there are some mathematical details that need to be worked out. These details don't change the overall intuitive picture, but they present yet again opportunities for confusion.

The first is the role of the Boltzmann constant $k_B$, which is a conversion factor between the Shannon entropy $S = - \sum p_i \log(p_i)$ and the Gibbs entropy $S_G = - k_B p_i \sum \log(p_i)$. The issue here is that temperature, internal energy and entropy are typically linked by the relationship $T = \frac{\partial U}{\partial S}$. Historically, temperature and internal energy are the quantities that were defined first. If we measure temperature in Kelvin and energy in Joules, the above relationship forces us to measure entropy in Joules/Kelvin. If we changed units, $k_B$ would change accordingly.

However, this is not the only way to define the relationships between those quantities. Another way is to have $\beta = \frac{1}{k_B T} = \frac{\partial S}{\partial U}$. In this setting, $U$ is measured in Joules, $S$ in nats (which are dimensionless), $\beta$ is measured in inverse Joules. Temperature is still measured in Kelvin, but we do not regard it as a primary objects. We find this approach, sometimes called entropy first, prefereable for a number of reasons, expecially in statistical mechanics. First, entropy is invariant under coordinate transformations and therefore it is better to keep it a pure number and not mix it with units of energy, which is not invariant under boost.\footnote{This is essential if one ever hopes to put statistical mechanics in a relativistic setting.} Second, the maximization of entropy at equilibrium guaranties us that $S(U)$ is a single valued function (i.e. the maximum given the value of $U$) and $\frac{\partial S}{\partial U}$ is therefore well defined. On the other hand, $U(S)$ is not in general single value (e.g. in the Ising model). This means that temperature can be infinite where the curve inverts, where small changes of energy do not change the entropy, and can be negative in the other region. That is, the infinitely cold system is temperature zero, which cannot be achieved, as it warms up temperature rises until it become infinite, and as the system becomes even hotter the temperature is negative and increase toward zero, which would correspond to an infinitely hot system. On the other hand, $\beta$ is always finite and cannot be infinite. An infinitely cold system, then would correspond to positive infinite $\beta$, as the system warms up beta decreases, passes from zero, becomes negative and an infinitely hot system corresponds to minus infinite $\beta$: this is a much better property for a physically meaningful quantity. Third, when setting up entropy maximization, $\beta$ is quite naturally the Lagrange multiplier associated with energy conservation, therefore mathematically it has a much clearer role. Lastly, one can proceed in both thermodynamic and statistical mechanics with just $\beta$, and convert to temperature only when we need a value on the Kelvin scale. For these reasons, it seems to us $\beta$ plays the more fundamental role and $k_B$ is the conversion coefficient to go from $\beta$ to temperature measured in Kelvin. Adding and removing the Boltzmann constant from the entropy expression, then, is equivalent to shift from an energy/temperature first approach to an entropy/$\beta$ first approach and should be not seen as physically significant.

The other point of possible confusion is that in statistical mechanics there are actually two distributions. A single microstate is itself a distribution of particles over their particle states. There are actually $N$ particles existing at the same time, each with its own particle state. And a macrostate is a probability distribution over microstates. It is important to keep these two distributions straight, especially because a microstate can be represented in two ways.

>>>>

\title{Universal Characterization of Physical Variability by \\Shannon's $-\sum_ip_i\,\textrm{ln}\,p_i$ Formula}

\author{
	Julian Barbour and 
	Gabriele Carcassi
}

@@

A world without any variety is barely conceivable. It certainly would not admit scientific investigation. One of us recalls but has been unable to relocate a sentence in one of Leibniz's many writings to the effect that reality \emph{is} variety. Whether he made the statement or not, it is consistent with his view that any contingent thing is defined by a collection of attributes, which necessarily must be distinguishable
at some epistemic level. Paragraph 58 of Leibniz's \emph{Mondadology} asserts that the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''.

Without committing ourselves to Leibniz's overarching principle, our aim in this paper is to draw attention to the specific nature of the variety that can be observed in the universe and the unique suitability of Shannon's entropy (or information) formula $-\sum_ip_i\,\textrm{ln}\,p_i$ for the characterization of the variety. Indeed, as we will show in the next section, opportunities for application of Shannon's formula are ubiquitous both on the earth and in the universe at large. PRECISELY BECAUSE OF THIS GENERALITY WE WILL OFTEN USE THE EXPRESSION SHANNON `FORMULA' RATHER THAN `ENTROPY' OR `INFORMATION'.

We also wish to comment on the various interpretations and uses of Shannon's formula and attempt to answer this question: what does the formula represent? It is well known that when Shannon recognized the importance of his formula in communication theory he sought von Neumann's advice on the name he should give to the expression he had found. He was told to call it \emph{entropy}, first, because Botlzmann had used an identical formula to define entropy in statistical mechanics and, second, nobody understood entropy and that would provide protection from criticism [exact words to be found]. We believe von Neumann's advice was not totally helpful and has given rise to confusion. Indeed, it is often claimed that Shannon's formula is epistemic in nature and represents ignorance. There certainly are situations in which that is true but equally there are others in which the use of the formula is very appropriate and no ignorance is involved at all. Examples will be given in the next section.

This leads us to propose a different characterization of Shannon's formula that is both more general and more appropriate. Indeed, we will argue that the formula is a measure of the \emph{variability within a distribution of things of a given kind}. Moreover, the formula has a very wide range of applicability because the universe contains many different kinds of things that, through having common attributes, can be recognized as belonging together despite the fact that their attributes are expressed to different degrees or in different ways. To give a first example, the girls in a school can differ in the length and colour of their hair. More examples will follow in the next section. In this paper we do not ask why the universe sustains the variety that it does. We take it as a given.

NEW PARAGRAPH

We conclude this section with a word on terminology. In mathematics, Cantor defined a set as ``a collection into a whole $M$ of definite and separate objects''. He called the objects \emph{elements}. We will also use the word `elements' to denote the things we consider but subject to the proviso that they have a distinguishing feature (or features) that allows us to say that \emph{they are of the same kind}. Thus, we consider sets of a restricted type. The point is this. To constitute a set in accordance with Cantor's definition, it is only necessary for its elements to be \emph{definite and separate} objects. This leaves total freedom as to how such objects in the universe are to be parcelled up into specific sets. For example, a set might consist of Mars and the word `go'. We think such latitude misses one of the most significant aspects of the universe, namely, that the things within it can be distinguished by attributes that are much more specific than the two required by Cantor. Our field of investigation goes beyond mere number, which washes out all attributes apart from the two, of being definite and separate, that allow us to count. They were the only attributes that Cantor needed; we need more.

\vspace{.5in}

The general setting is the following. We have a set of elements $E = \{e_\alpha\}_{\alpha=1}^N$. In accordance with the comments made in our introduction, they can for example represent the galaxies in the Laniakea Supercluster, the animals in the Gal\'{a}pagos islands, the molecules in a box of gas given a particular macrostate, the different outcomes of spin measurement for a given quantum state, or the words in the complete works of Shakespeare. The choice of the particular set is driven by the interest and practices of scientists and their fields of study and therefore it is, in this sense, discretional. Yet, it must be objective in at least one respect: once a choice is made the actual elements are the same for everybody. Once we arbitrarily decided we will study the galaxies in the Laniakea Supercluster, it is a matter of fact that the Milky Way will be included. This also means that, whatever properties those elements will have, they will represent a matter of fact about that set. In turn, this will make some choices arguably more appropriate or interesting than others. For example, in taxonomy the set of all animals that have feathers may be more interesting than the set of all animals that can fly: the first includes only birds while the second includes most birds, a few mammals and a lot of insects.

Once a set of elements is chosen, we select a property or a set of properties we want to use to characterize the elements. That is, we have a set of possible values $Q=\{q_i\}_{i=1}^I$ and a map $q : E \to Q$ that associates a value for each element. These can represent the galaxy types, the genus of the animal, the possible states of the molecules, the possible values for spin or the spelling of words. This will give us a sequence $\{q(e_\alpha)\}_{\alpha=1}^N$ of the descriptions that are associated with each element in the set. Again, the choice of properties is ultimately discretional, grounded on what particular aspects of the elements we are studying in a given case. Yet, like before, once a choice is made the value for each element is objectively defined. Once we arbitrarily decided we are studying galaxy types it is a matter of fact that the Milky Way is a barred spiral galaxy. Again, one may find that some choices are arguably more appropriate or interesting than others.

Once we have constructed our distribution, we will want to construct an indicator $S$ that quantifies how much variability the elements exhibit within the distributions. That is, we want to quantify the degree of diversity that the values can have within the distribution. To that end, we want to define some suitable requirements for $S$.

We may be tempted to use standard statistical quantities, like the range or the variance, but this is not possible. First, if the values associated with the bins are non-numerical (e.g. types of galaxies, words), statistical quantities are not well defined. Second, relabeling the values (e.g. switching names, switching units, non-linearly changing coordinates) does not have an impact on variability of the elements, while statistical quantities will in general be affected. This tells us that our indicator cannot depend on $Q$ itself, but only on the weights $p_i$. That is, we require $H=H(p_i)$.\footnote{Note that for continuous quantities the weights are densities and are affected by the choice of $Q$: the unit is required to specify the numeric value (e.g. $1 \% / m$) and this will change under unit transformation (e.g. $1000 \% / Km$). Therefore, as the weights $p_i$ themselves depend on the unit, the indicator $H(p_i)$ will in general depend on the choice of $Q$. Though this is source of additional confusion, for the purpose of defining a measure of variability it does not change things conceptually. We will discuss the problem in connection to entropies in thermodynamics and statistical mechanics, since coordinate invariance is paramount in physics.}

We  expect small changes in a distribution -- small changes of the weights -- to produce small changes in variability; we therefore require $H$ to be a continuous function of the $p_i$. We also expect that as the number of values found within the distribution increases, so will the variability. Therefore if we have a uniform distribution over $I$ cases, so that $p_i = 1/I$, we require $H$ to be monotonically increasing with $I$.

To sum up, we have the following three requirements:
\begin{enumerate}
\item $H$ depends only on $p_i$ and it does so continuously
\item If $p_i=1/I$ then $H$ is a monotonically increasing function of $I$
\item Let $p_i$ and $q_j$ be the weights for two distributions respectively over $I$ and $J$ bins. Let $r_k$ be the distribution over $K=I+J-1$ bins constructed by expanding the $a$-th bin of the first distribution using the second distribution. More specifically, let $1 \leq a \leq I$, then $r_k = \{p_1, p_2, ..., p_{a-1}, p_{a}q_1, p_{a}q_2, ..., p_{a}q_J, p_{a+1}, ..., p_I \}$. Then $H(r_k) = H(p_i) + p_{a} H(q_j)$.
\end{enumerate}

The Shannon entropy, then, is a clear and important attribute of a distribution: it quantifies the variability of the elements within the distribution. It quantifies the variety of values one finds within the distribution. It is true that the choice of the elements and binning are not fixed, it is true that the meaning of the weights depends on what the distribution is describing, and therefore that the precise meaning of this variability is context dependent. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope a probability and so on. And like a real number represents a quantity, transcending what is physically quantified, the Shannon entropy represents the variability of the elements\footnote{We again stress that out elements are physical and therefore posses some property beyond the bare `definite and separate' of Cantor's definition, which was perfectly suited to study numbers. Without these extra attributes, which are `washed out' by said definition, we would not be able to create a distribution in the first place.} in a distribution, transcending what the distribution physically signifies.

The Shannon entropy may represent knowledge in some cases, but not in general. Consider the following two cases:
\begin{enumerate}
\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you didn't.
\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you won half a million dollars.
\end{enumerate}
The distribution in both cases is the same, two bins 50\% chance each, and so is the Shannon entropy. Yet, you know more in the second case: you know you have won at least half a million dollars.

Unfortunately, entropy in general is often associated with vague characterizations like the the two presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending by the authors[CITATION NEEDED], which leads to confusion and misunderstanding.  The characterization of the Shannon variability measure we have given, on the other hand, applies to all cases and leads naturally to the assumptions required to rederive it. OUR CHARACTERIZATION IS therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos island, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability we find in his vocabulary? All these questions are interesting, all these questions are profound and the answer to all these questions are expressed through just one thing: the Shannon variability measure.

If the Shannon expression is a measure of variability, why is it connected to information? What is it measuring and in what units?

We can gain insight with the following example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands grouped by their respective species. Suppose we pick a specific animal from the set and you want to know its species. Suppose the only way for you to get that information is to ask a series of questions with only two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

Not all questions will be able to extract the same amount of information. Some questions, like ``is it an animal?" would be redundant. Others, like ``is it an American Flamingo (Phoenicopterus ruber)?", would give us a lot of information in the positive case but little in the negative case. However, it is clear that there is always a minimum number of questions that must be asked to get to the answer. Given that there are more than two species, a single question cannot be enough. It should however be intuitively clear that to a greater variability within the distribution will correspond a greater number of questions you must ask. That is exactly what the Shannon variability quantifies: the minimum average number of questions one has to ask to identify a value in the distribution. It gives us the number of questions for an ideal strategy for our game of Twenty Questions.

The Shannon entropy of the distribution, then, has nothing to do with the information the distribution itself holds. The distribution is not what is being encoded. The entropy is the information needed to go from the distribution, which is given, to an individual element. It's really the information gap from the population to an element. That is why some people say the entropy is ``lack of information", because it's the information the distribution does not have. But, again, this is just missing the point of what is being described. If one is not interested in identifying elements there is no ``missing information".

As we saw in the previous sections, Shannon variability is a very general concept that applies to any type of distribution, no matter what the problem domain. It does not require to define states, processes, reversibility and therefore it is conceptually different and independent from the entropy in thermodynamic and statistical mechanics which requires them.\footnote{For this reason it is unfortunate that it was given the same name.} It is only when variability is appropriately applied to this particular setting that, though not apparent at first, a connection can be made. To understand this link we need to answer two basic questions. First, what does the Shannon variability represent in the context of statistical mechanics and thermodynamics? In particular, why is it a state variable and what property of the system is it describing? Second, why does the Shannon variability satisfy the same properties as the entropy? Namely, why does it increase during irreversible evolution, why is it maximized at equilibrium and why is it additive for independent systems?

The first question is straightforward to answer. The macroscopic variables of a system (e.g. volume, pressure, energy) do not give a full description of its parts. If we fix a macrostate and then measure the state of all its microscopic constituents, we will not get the same answer every time. This true in the case of non equilibrium (e.g. we suddenly compress a gas pushing a piston) and of equilibrium (e.g. we give enough time to equilibrate). The choice of a macrostate, then, identifies a set of possible microstates that are compatible with it. It is then valid to ask the question: what is the variability of microstates one can find within a given macrostate? This is answered by the Shannon variability, which gives us the number of bits required to identify a microstate given the macrostate. Since the macrostates fully defines the distribution of possible microstates, this distribution does not depend on the particular way the macrostate came to pass. It is path independent: it is a state variable. In short, as a macrostate is a distribution over microstates, it will admit as a state variable the Shannon variability, regardless of whether we know the actual distribution.

Note, though, that the thermodyanmic entropy is defined differently. It is defined over a succession of equilibrium states by noting that the integral of the heat exchanged divided by the temperature is path independent. One then chooses a particular state $A$, assigns it an arbitrary value $S_A$ and the entropy for all other states is given by the integral $S(B) = S_A + \int_A^B \frac{dQ}{T}$. At this point, it is not apparent why these two quantities defined independently would be related. What we can say, though, is that all quantities that increase over irreversible processes, are maximized at equilibrium and that are additive for indipendent systems are precisely the quantities that are linear functions of entropy.[CITE AXIOMATIC WORK] A linear function of entropy, though, simply means a redefinition of $S_A$ and a change of units for either $dQ$ or $T$. Therefore if we can show that the Shannon variability possesses those three properties than it is equivalent to the entropy up to a change in units, which brings us to the second question.

Suppose we have an irreversible process, which means we can start from many out-of-equilibrium states which end into a single equilibrium state. During the whole process, some of the macro variables are going to stay constant as they characterize the constraint the system is subjected during the evolution. For example, if the walls are fixed, the gas will have a set volume; if the walls are insulated, then the average energy will be conserved.\footnote{It can also happen that the irreversible process does have a single possible equilibrium, as in the case of spontaneous symmetry breaking. New variables will identify the equilibrium which we can think of new constraints appeared once the initially symmetry is broken.} As different out-of-equilibrium macrostates merge during the evolution, they merge the possible initial conditions that lead to the same macrostate. The variability of microstates within the initial out-of-equilibrium macrostate, then, must be smaller than the variability of the microstates within the final equilibrium state. The amount of information needed to identify a microstate at equilibrium allows us to identify the initial microstate across multiple out-of-equilibrium states and therefore is greater than the information needed to identity a microstate in an out-of-equilibrium macrostate. Irreversible processes increase the Shannon variability within the macrostate, which corresponds to a greater uniformity of the macrostate itself since uniform distributions are the ones that maximize the variability.

To understand why this is maximized at equilibrium, suppose it weren't. That is, suppose there was another macrostate which satisfied the same constraints, but for which the Shannon variability was greater. If we prepared such a state and let it equilibrate, it could not evolve to the equilibrium because that has lower Shannon variability. Its equilibrium, then, would be different from the previous one, but this cannot be since we said that those constraints would lead to a single equilibrium state. Therefore the Shannon entropy is maximized for an equilibrium state.

Now suppose we have two independent systems, like two volumes of gasses. As they are independent, picking the microstate of one system does not constrain the microstate of the other. That is, the distribution over microstates are statistically independent. This means that the statistical distribution is the product of the join distributions and, in this case, the Shannon entropy is additive.

The Shannon variability, then, has all the right properties to establish a conceptually meaningful connection with the thermodynamic/statistical mechanics entropy. It is indeed an independent concept but when applied to macrostates yields a state variable with the right conceptual characteristics. There are still some mathematical details that need to be worked out to have a complete picture. These details don't change the overall intuitive picture, but they present yet again opportunities for confusion.

The first is the role of the Boltzmann constant $k_B$, which is a conversion factor between the Shannon variability. $S = - \sum p_i \log(p_i)$ and the Gibbs entropy $S_G = - k_B p_i \sum \log(p_i)$. This conversion constant should not surprise us, since we only established a linear relationship between variability and entropy. The question is whether this conversion itself is has a physical meaning, or it is also simply due to a choice of units. The standard relationship between temperature, internal energy and entropy is $T = \frac{\partial U}{\partial S}$. Historically, temperature and internal energy are the quantities that were defined first. If we measure temperature in Kelvin and energy in Joules, the above relationship forces us to measure entropy in Joules/Kelvin. If we changed units, $k_B$ would change accordingly.

However, this is not the only way to define the relationships between those quantities. Another way is to have $\beta = \frac{1}{k_B T} = \frac{\partial S}{\partial U}$. In this setting, $U$ is measured in Joules, $S$ in nats (which are dimensionless), $\beta$ is measured in inverse Joules. Temperature is still measured in Kelvin, but we do not regard it as a primary object. We find this approach, sometimes called entropy first, preferable for a number of reasons, especially in statistical mechanics. First, entropy is invariant under coordinate transformations and therefore it is better to keep it a pure number and not mix it with units of energy, which is not invariant under boost.\footnote{This is essential if one ever hopes to put statistical mechanics in a relativistic setting.} Second, the maximization of entropy at equilibrium guaranties us that $S(U)$ is a single valued function (i.e. the maximum given the value of $U$) and $\frac{\partial S}{\partial U}$ is therefore well defined. On the other hand, $U(S)$ is not in general single valued (e.g. in the Ising model). This means that temperature can be infinite where the curve inverts, where small changes of energy do not change the entropy, and can be negative in the other region. That is, the infinitely cold system is temperature zero, which cannot be achieved, as it warms up temperature rises until it becomes infinite, and as the system becomes even hotter the temperature is negative and increase toward zero, which would correspond to an infinitely hot system. On the other hand, $\beta$ is always finite and cannot be infinite. An infinitely cold system, then would correspond to positive infinite $\beta$, as the system warms up beta decreases, passes from zero, becomes negative and an infinitely hot system corresponds to minus infinite $\beta$: this is a much better behavior for a physically meaningful quantity. Third, when setting up entropy maximization, $\beta$ is quite naturally the Lagrange multiplier associated with energy conservation, therefore mathematically it has a much clearer role. Lastly, one can proceed in both thermodynamic and statistical mechanics with just $\beta$, and convert to temperature only when we need a value on the Kelvin scale. For these reasons, it seems to us $\beta$ plays the more fundamental role and $k_B$ is the conversion coefficient to go from $\beta$ to temperature measured in Kelvin. Adding and removing the Boltzmann constant from the entropy expression, then, is equivalent to shift from an energy/temperature first approach to an entropy/$\beta$ first approach and should be not seen as physically significant. It is really just fixing a unit.

The other point of possible confusion is that in statistical mechanics there are actually two distributions. A single microstate is itself a distribution of particles over their particle states. There are actually $N$ particles existing at the same time, each with its own particle state. And a macrostate is a probability distribution over microstates. It is important to distinguish between these two distributions, especially because a microstate can be represented in two ways.

The first way is as a distribution over a single copy of phase space, sometimes referred to as $\mu$-space. In this case the distribution represents the fraction of particles that are in that state. The Maxwell-Boltzmann distribution $\rho(q^i, p_i) = \left(\frac{\beta m}{2\pi} \right)^{3/2}e^{-\beta \frac{p_ip^i}{2m}}$ is expressed in this space. On $\mu$-space, the Shannon variability $-\int\rho \log \rho$ does not coincide with the thermodynamic entropy for a simple reason: it does not depend linearly on the number of particles. It is not an extensive quantity. The reason is straightforward: the Shannon entropy works on normalized distributions and gives us the variability of a single element. In this case, the system is composed by $N$ elements, not one. One can use the expression $-N\int\rho \log \rho$ as this represents the variability of $N$ elements.

We want to stress that a distribution on $\mu$-space can only either describe a single microstate where $N$ is very large and therefore the distribution can be assumed to be continuous, or a probability distribution for a $N$ identically and independently prepared particles. It one wanted to describe a macrostate, one would have to work with a probability distribution (i.e. the weights represents probabilities) over the space of particle distributions (i.e. the weights represent fractions of total particles) over $\mu$-space or one would obtain erroneous results.\footnote{This happens, for example, in [Jaynes] where he creates the distribution over $\mu$-space by averaging over all microstates. This essentially blends together the probabilities and particle counts losing most of the actual statistical information. Naturally, the entropy thus computed is wrong and found to not work.}

Alternatively, a microstate can be represented as a point over $N$ copies of phase space, sometimes referred to as $\Gamma$-space. It is on this space that the Gibbs entropy is defined and where thermodynamic entropy happens to coincide with the Shannon entropy. Since this is a probability distribution, the microstate corresponds to a single element within the distribution defined by the macrostate. Yet, one has to be careful of two things. The first is that permutations of particles within $\Gamma$-space lead to different points which should not lead to different microstates. This leads to the widely known problem of overcounting which needs to be addressed in the standard way. The second problem is that we should not think of the particles as literally pointlike: they would correspond to delta Dirac distributions over phase-space which have minus infinite Shannon entropy. It is more appropriate, both mathematically and conceptually, to think of indistinguishable particles as identical distributions over $\mu$-space, each contributing the same amount of Shannon entropy.\footnote{Setting to $-\log h$ the entropy corresponding to each degree of freedom of these single particle distributions is a natural way to incorporate the effects of the uncertainty principle.}

@@
